`r opts_chunk$set(cache=TRUE, fig.width=12)`
<link href="http://joey711.github.com/phyloseq/markdown.css" rel="stylesheet"></link>

The phyloseq Extensions Index
========================================================

This page is automagically updated when I do a periodic full rebuild of the phyloseq extensions documentation. The date of this particular re-build is `r date()`.

Don't forget to checkout [the phyloseq demo repository](http://joey711.github.io/phyloseq-demo/) for other tutorials; some more in-depth or lengthy than are maintained here, where the focus is documenting phyloseq extension functionality rather than demonstrating use cases with new/large datasets.

---
## Available Extensions

```{r find-parse-tutorials, echo=FALSE, include=FALSE}
files = list.files(pattern="\\.[Rr]md")
stems = gsub("\\.[Rr]md", "", files)
# Remove the index line
stems = stems[stems != "extensions-index"]
links = paste(stems, ".html", sep="")
mdlines = paste("### [", stems, "](", links, ")", sep="")
```

```{r write-links, results='asis', echo=FALSE}
cat(mdlines, sep="\n\n")
```


---
## The R Markdown Source Code
For more advanced users interested in seeing and/or modifying the [R markdown](http://www.rstudio.com/ide/docs/r_markdown) source code from which these reproducible, web browser-friendly tutorials were created.

```{r rmd-links, results='asis', echo=FALSE}
rmdlines = paste("### [", files, "](", files, ")", sep="")
cat(rmdlines, sep="\n\n")
```